---
title: "How Report Card Scores Are Calculated"
format:
  html:
    embed-resources: true
    html-table-processing: none
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    css: ../styles.css
lightbox: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tbeptools)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(gt)
library(here)

source(here('R/funcs.R'))
load(here('data/trndat.rda'))

# example year and groups used throughout
ex_yr  <- 2025
ex_grps <- sort(unique(trndat$grpact[trndat$yr == ex_yr]))

# pre-compute objects reused across sections
truespp  <- truespp_fun(trndat, ex_yr)
truvar   <- truvar_fun(trndat, ex_yr)
cal      <- calibrate_scr_fun(trndat)
allgrpscr <- allgrpscr_fun(trndat, ex_yr, truvar, cal = cal)

# non sav species
non <- '^AA|^DA|^DB|^DG|^DR|^CYAN'

# # pick the median-scoring group as the "worked example" group
# ex_grp <- allgrpscr |>
#   dplyr::arrange(as.numeric(factor(Total,
#     levels = c('A','A-','B+','B','B-','C+','C','C-','D+','D')))) |>
#   dplyr::slice(ceiling(n() / 2)) |>
#   dplyr::pull(grpact)

# example group
ex_grp <- trndat |> 
  dplyr::filter(yr == ex_yr) |>
  dplyr::filter(grp == 'C') |> 
  dplyr::pull(grpact) |> 
  unique()

ex_grp_label <- trndat |> 
  dplyr::filter(grpact == ex_grp) |>
  dplyr::pull(grp) |>
  unique()

evalgrp <- evalgrp_fun(trndat, ex_yr, ex_grp, truvar)
```

## Overview

Seagrass transects are monitored in Tampa Bay each year by multiple resource management agencies.  To support these efforts, the Tampa Bay Estuary Program coordinates an annual intercalibration training prior to the field season.  This intercalibration effort ensures that all participating groups are using consistent methods and interpretations. Seagrasses are independently assessed by each group at the same locations and the results are compared using report cards. This document describes how the report cards are prepared and what the scores mean.  The intent is to provide each group with a general indication of how closely their measurements align with the consensus of all groups and to identify areas for improvement.

Because there is no single external ground truth, scores are based on how consistently groups agree with each other. A group that reports values close to the cross-group average earns a high score and one that deviates substantially earns a lower score.

Scores are calculated for three field measurements:

- **Abundance**: Braun-Blanquet cover category (0, 0.1, 0.5, 1, 2, 3, 4, 5)
- **Blade Length**: average blade length in cm
- **Short Shoot Density**: shoots per m²

Taxa (seagrass and macroalgae, if present) observed relative to other groups is also considered. These three metric scores are averaged into an overall total score, which is then converted to a letter grade.

Note that "transect" is used herein to describe the intercalibration sites, which are quadrats at fixed locations.  For the actual transect sampling, sites are set meter marks along a transect line.

## Step 1: The Consensus Species List

Not every species recorded across all groups counts toward scoring. A species at a given transect (site/quadrat) is considered present if at least two distinct groups reported it with non-zero abundance. This removes likely misidentifications while still being inclusive, i.e., a species does not need to be unanimous, just corroborated.

```{r}
#| label: tbl-truespp
#| tbl-cap: !expr paste("Consensus species at each transect in", ex_yr, "- species reported by at least two groups.")

truespp |>
  dplyr::count(Site, name = 'n_species') |>
  dplyr::mutate(
    Species = purrr::map_chr(Site, function(s){
      truespp |>
        dplyr::filter(Site == s) |>
        dplyr::pull(Species) |>
        sort() |>
        paste(collapse = ', ')
    }), 
    Site = as.numeric(Site)
  ) |>
  dplyr::rename(Transect = Site, `Species count` = n_species) |>
  dplyr::arrange(Transect) |> 
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(style = 'italic'),
    locations = gt::cells_body(columns = Species)
  ) |>
  gt::cols_align('left') |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

The consensus list is created per transect, so a species may be on the list at one transect but not another.

## Step 2: True Values

For each consensus species at each transect, true values are calculated as the cross-group average:

- **Abundance**: group Braun-Blanquet values (0, 0.1, 0.5, 1, 2, 3, 4, 5) are averaged directly across groups, then snapped to the nearest valid BB value (no coverage, solitary, few, <5%, 5-25%, 26-50%, 51-75%, 76-100%).
- **Blade Length** and **Short Shoot Density**: simple means across groups.

Only consensus species enter this calculation. This prevents a misidentified species (recorded by only one group) from distorting the true values for everyone else.

```{r}
#| label: tbl-truvar
#| tbl-cap: !expr paste('Example "true" values at one transect in', ex_yr, 'determined by cross-group averages for consensus species.')

truvar |>
  dplyr::filter(Site == '8') |> #min(as.numeric(Site))) |>
  tidyr::pivot_wider(names_from = var, values_from = truval) |>
  dplyr::mutate(
    `Abundance (0, 0.5, 1, 2, 3, 4, 5)` = as.character(Abundance),
    `Blade Length (cm)` = ifelse(grepl(non, Species), 'NA', round(`Blade Length`, 1)),
    `Short Shoot Density (per m²)` = ifelse(grepl(non, Species), 'NA', round(`Short Shoot Density`, 1))
  ) |>
  dplyr::select(Transect = Site, Species, `Abundance (0, 0.5, 1, 2, 3, 4, 5)`, `Blade Length (cm)`, `Short Shoot Density (per m²)`) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(style = 'italic'),
    locations = gt::cells_body(columns = Species)
  ) |>
  gt::cols_align('left') |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

## Step 3: Group Deviations and Species ID Penalties

For each group, reported values are compared to the true values across all consensus species and transects. This comparison uses a full join, which reveals two types of species identification errors:

**Missed species**: a species is present but the group did not record it. The group's abundance for that species is treated as level 1 ("no coverage", numeric conversion for BB value 0) when computing the deviation, applying a slight penalty.

**False positives**: the group recorded a species that is not on the consensus list (i.e., no other group confirmed it). The true abundance is treated as level 1, again applying a slight penalty.

These penalties only affect the abundance score. Blade Length and Short Shoot Density cannot be meaningfully penalised for species that were not found.

```{r}
#| label: tbl-evalgrp
#| tbl-cap: !expr paste0('Reported vs. true values for group "', ex_grp_label, '" at one transect in ', ex_yr, '.')

evalgrp |>
  dplyr::filter(Site == '8') |>
  dplyr::select(
    Transect = Site, Species,
    `Abundance reported` = `Abundance aveval`,
    `Abundance true`     = `Abundance truval`,
    `Blade Length reported` = `Blade Length aveval`,
    `Blade Length true`     = `Blade Length truval`, 
    `Short Shoot Density reported` = `Short Shoot Density aveval`,
    `Short Shoot Density true`     = `Short Shoot Density truval`
  ) |>
  dplyr::mutate(dplyr::across(`Abundance reported`:`Abundance true`, ~ ifelse(is.na(.), '—', as.character(.)))) |>
  dplyr::mutate(dplyr::across(`Blade Length reported`:`Short Shoot Density true`, ~ ifelse(grepl(non, Species), 'NA', as.character(.)))) |>
  dplyr::arrange(Species) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(style = 'italic'),
    locations = gt::cells_body(columns = Species)
  ) |>
  gt::cols_align('left') |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

## Step 4: Metric Scores

For each metric (abundance, blade length, short shoot density), deviations from the true value are summarised per species across transects, then combined into a single number per group per metric. The combination uses a weighted mean of absolute differences, where the weight for each species is the inverse of the standard deviation of the true values across transects:

$$
\text{metric score}_{\text{raw}} = \sum_{\text{species}} \frac{|\bar{d}_s|}{1 + \sigma_s}
$$

where $\bar{d}_s$ is the mean absolute deviation for species $s$ and $\sigma_s$ is the standard deviation of the true values for that species across transects. Species where the true value varies a lot across transects (high $\sigma_s$) contribute less to the final score, since agreement there is inherently harder to achieve.

```{r}
#| label: tbl-sppdiff
#| tbl-cap: !expr paste0("Per-species deviation summary of the abundance metric for group '", ex_grp_label, "' in ", ex_yr, ".")

sppdiff_fun(evalgrp, 'Abundance') |>
  dplyr::mutate(
    aveval  = factor(aveval  - 1, levels = 0:7,
                     labels = c('no coverage','solitary','few','<5%','5-25%','25-50%','51-75%','76-100%')),
    truval  = factor(truval  - 1, levels = 0:7,
                     labels = c('no coverage','solitary','few','<5%','5-25%','25-50%','51-75%','76-100%')),
    avediff = round(avediff, 2)
  ) |>
  dplyr::select(Species, `Reported (avg)` = aveval, `True (avg)` = truval,
                `Mean deviation` = avediff) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(style = 'italic'),
    locations = gt::cells_body(columns = Species)
  ) |>
  gt::cols_align('left') |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

## Step 5: Score Calibration

Raw metric scores are converted to a 0–100 scale. Without calibration, the best group in any year always maps to 100 and the worst always maps to 50, regardless of how closely groups agreed. This means a year where everyone performed very well would still produce a spread from A to D, which needs to be accounted for to avoid an unfair outcome.

To address this, the score floor (the minimum possible score) is raised in years when all groups agree closely with each other, and kept at 50 in years when disagreement is high.

### How calibration works

For each year and metric, we compute the within-year standard deviation of group deviations to assess the spread between groups. This is then expressed as a ratio to the historical mean spread across all training years. A ratio below 1 means groups agreed more closely than usual; a ratio above 1 means more disagreement than usual.

The score floor for a given year and metric is:

$$
\text{floor} = \max\!\left(50,\ 50 + \left(1 - \frac{\text{SD}_{\text{year}}}{\overline{\text{SD}}}\right) \times 25\right)
$$

where $\text{SD}_{\text{year}}$ is the within-year spread for that metric and year, $\overline{\text{SD}}$ is its historical mean, and 25 is a scaling constant that sets the maximum possible floor lift (grade-points). Choosing the scaling constant is subjective and determines the basis for how similar scores should be when deviation between groups in a year is tight.  Raising the value makes the scores more similar and lowering the value makes them more spread out. The maximum lift occurs when all groups agree perfectly (ratio = 0), giving a floor of $50 + 25 = 75$, which corresponds to a minimum grade of B-. A year at the historical average (ratio = 1) receives no adjustment and keeps a floor of 50. Loose years (ratio > 1) are capped at 50 so that no extra penalty is applied beyond the standard range.

```{r}
#| label: fig-calibration
#| fig-cap: "Within-year spread of group deviations for each metric and year. Bar colour shows the ratio of each year's spread to the historical mean (dashed line). Bar labels show the ratio and the resulting score floor. Blue (ratio < 1) years are tighter than average and receive a higher score floor."
#| fig-height: 5
#| fig-width: 9.5

yrs <- sort(unique(trndat$yr))

cal_lkp <- tibble::tibble(
  metric  = c('Abundance', 'Blade Length', 'Short Shoot Density'),
  mean_sd = c(cal$mean_sd$Abundance, cal$mean_sd$`Blade Length`,
              cal$mean_sd$`Short Shoot Density`)
)

yr_spreads <- purrr::map(yrs, function(yr){
  tv <- truvar_fun(trndat, yr)
  allgrpscr_fun(trndat, yr, tv, raw_diff = TRUE) |>
    dplyr::summarise(dplyr::across(Abundance:`Short Shoot Density`,
                                   ~ sd(.x, na.rm = TRUE))) |>
    dplyr::mutate(yr = yr)
}) |>
  dplyr::bind_rows() |>
  tidyr::pivot_longer(-yr, names_to = 'metric', values_to = 'sd_avediff') |>
  dplyr::left_join(cal_lkp, by = 'metric') |>
  dplyr::mutate(
    ratio     = sd_avediff / mean_sd,
    floor_adj = as.integer(pmax(50, 50 + (1 - ratio) * 25)),
    metric    = factor(metric, levels = c('Abundance', 'Blade Length', 'Short Shoot Density'))
  )

ggplot(yr_spreads, aes(x = factor(yr), y = sd_avediff, fill = ratio)) +
  geom_col(width = 0.7, colour = 'grey40', linewidth = 0.3) +
  geom_hline(aes(yintercept = mean_sd), linetype = 'dashed', linewidth = 0.6) +
  geom_text(aes(label = sprintf('ratio=%.2f\nfloor=%d', ratio, floor_adj)),
            vjust = -0.25, size = 2.5, lineheight = 0.9) +
  scale_fill_gradient2(low = '#2166ac', mid = '#f7f7f7', high = '#d73027',
                       midpoint = 1, name = 'Ratio') +
  scale_y_continuous(expand = expansion(mult = c(0, 0.28))) +
  facet_wrap(~ metric, scales = 'free_y', nrow = 1) +
  labs(x = NULL, y = 'Within-year SD of group deviations') +
  theme_bw(base_size = 10) +
  theme(legend.position = 'bottom',
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_rect(fill = '#004F7E'),
        strip.text = element_text(colour = 'white', face = 'bold'))
```

### Effect on scores: tight vs. loose year

The table below compares the calibrated score floor for each year and metric, illustrating how the floor shifts in tighter training years.

```{r}
#| label: tbl-floors
#| tbl-cap: "Calibrated score floor by year and metric. Years with consistently tight agreement receive a higher floor, raising the minimum grade for all groups in that year."

yr_spreads |>
  dplyr::select(yr, metric, floor_adj) |>
  tidyr::pivot_wider(names_from = metric, values_from = floor_adj) |>
  dplyr::rename(Year = yr) |>
  gt::gt() |>
  gt::data_color(
    columns = c(Abundance, `Blade Length`, `Short Shoot Density`),
    fn = scales::col_numeric(
      palette = c('#c6dbef', '#c7e9c0'),
      domain = c(50, 75)
    )
  ) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::cols_align('center', columns = -Year) |>
  gt::tab_options(
    row.striping.include_table_body = FALSE,
    row.striping.background_color = 'white',
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

## Step 6: Letter Grades

After calibration, each group's numeric score for each metric falls on a 0–100 scale. These are mapped to letter grades using fixed thresholds:

```{r}
#| label: tbl-grades
#| tbl-cap: "Letter grade thresholds. Scores at or above the lower bound but below the upper bound receive the listed grade."

tibble::tibble(
  Grade        = c('A', 'A-', 'B+', 'B', 'B-', 'C+', 'C', 'C-', 'D+', 'D'),
  `Score range` = c('95 – 100', '90 – 94', '85 – 89', '80 – 84',
                    '75 – 79', '70 – 74', '65 – 69', '60 – 64', '55 – 59', 'below 55')
) |>
  gt::gt() |>
  gt::cols_align('center') |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

The Total score is the unweighted average of the Abundance, Blade Length, and Short Shoot Density numeric scores, then converted to a letter grade using the same thresholds.

## Worked Example

```{r}
ex_grpscr <- allgrpscr_fun(trndat, ex_yr, truvar, cal = cal) |>
  dplyr::filter(grpact == ex_grp)

ex_grpscr_raw <- allgrpscr_fun(trndat, ex_yr, truvar, cal = cal, raw = TRUE) |>
  dplyr::filter(grpact == ex_grp)
```

The following walks through the full scoring pipeline for group `r ex_grp_label` in `r ex_yr`.

### Raw deviations

```{r}
#| label: tbl-example-sppdiff
#| tbl-cap: !expr paste0("Per-species deviations across all metrics for group ", ex_grp_label, " in ", ex_yr, ".")

purrr::map(c('Abundance', 'Blade Length', 'Short Shoot Density'), function(vr){
  sppdiff_fun(evalgrp, vr) |>
    dplyr::mutate(metric = vr, avediff = round(avediff, 2)) |>
    dplyr::select(metric, Species, `Reported avg` = aveval, `True avg` = truval,
                  `Mean deviation` = avediff)
}) |>
  dplyr::bind_rows() |>
  gt::gt(groupname_col = 'metric') |>
  gt::tab_style(
    style = gt::cell_text(style = 'italic'),
    locations = gt::cells_body(columns = Species)
  ) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#004F7E'),
                 gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_row_groups()
  ) |>
  gt::cols_align('left') |>
  gt::tab_options(
    table.border.top.color = '#004F7E',
    table.border.bottom.color = '#004F7E',
    column_labels.border.bottom.color = '#004F7E'
  )
```

### Scores

```{r}
#| label: tbl-example-scores
#| tbl-cap: !expr paste0("Final scores for group ", ex_grp_label, " in ", ex_yr, " showing the numeric score (calibrated) and letter grade.")

dplyr::bind_rows(
  ex_grpscr_raw |>
    tidyr::pivot_longer(-grpact, names_to = 'Metric', values_to = 'value') |>
    dplyr::mutate(type = 'Numeric score', value = as.character(round(as.numeric(value), 1))),
  ex_grpscr |>
    tidyr::pivot_longer(-grpact, names_to = 'Metric', values_to = 'value') |>
    dplyr::mutate(type = 'Letter grade')
) |>
  dplyr::select(-grpact) |>
  tidyr::pivot_wider(names_from = type, values_from = value) |>
  dplyr::mutate(
    Metric = factor(Metric, levels = c('Abundance', 'Blade Length', 'Short Shoot Density', 'Total'))
  ) |>
  dplyr::arrange(Metric) |>
  gt::gt() |>
  gt::cols_align('center', columns = -Metric) |>
  gt::tab_style(
    style = gt::cell_text(weight = 'bold'),
    locations = gt::cells_body(rows = Metric == 'Total')
  ) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

### How the calibration affected this group's scores

```{r}
#| label: tbl-example-cal
#| tbl-cap: !expr paste0("Effect of calibration on scores for group ", ex_grp_label, " in ", ex_yr, ". The floor is the lowest possible score any group could receive in this year for each metric.")

yr_spreads_wide <- yr_spreads |>
  dplyr::filter(yr == ex_yr) |>
  dplyr::select(metric, ratio, floor_adj)

ex_grpscr_nocal_raw <- allgrpscr_fun(trndat, ex_yr, truvar, raw = TRUE) |>
  dplyr::filter(grpact == ex_grp) |>
  tidyr::pivot_longer(-grpact, names_to = 'metric', values_to = 'score_before') |>
  dplyr::select(-grpact)

ex_grpscr_cal_raw <- ex_grpscr_raw |>
  tidyr::pivot_longer(-grpact, names_to = 'metric', values_to = 'score_after') |>
  dplyr::select(-grpact)

dplyr::left_join(ex_grpscr_nocal_raw, ex_grpscr_cal_raw, by = 'metric') |>
  dplyr::left_join(yr_spreads_wide, by = 'metric') |>
  dplyr::filter(metric != 'Total') |>
  dplyr::mutate(
    dplyr::across(c(score_before, score_after), ~ round(as.numeric(.), 1)),
    ratio = round(ratio, 2),
    floor_adj = as.integer(floor_adj)
  ) |>
  dplyr::rename(
    Metric = metric,
    `Score without calibration` = score_before,
    `Score with calibration` = score_after,
    `Ratio (within-year spread / historical mean)` = ratio,
    `Score floor` = floor_adj
  ) |>
  dplyr::mutate(
    Metric = factor(Metric, levels = c('Abundance', 'Blade Length', 'Short Shoot Density'))
  ) |>
  dplyr::arrange(Metric) |>
  gt::gt() |>
  gt::cols_align('center', columns = -Metric) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = '#2166ac'), gt::cell_text(color = 'white', weight = 'bold')),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.border.top.color = '#2166ac',
    table.border.bottom.color = '#2166ac',
    column_labels.border.bottom.color = '#2166ac'
  )
```

A ratio below 1 (tighter than average cohort) raises the floor for all groups, including this one. A floor of 50 means no calibration adjustment was applied.
